pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-08-04 bmc_article.tex
changed links to supp figures

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figures 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 16 and 24). This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figures 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figures 33-34). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

The set of alternatives found for the overall run (Figure 8 and Supplementary Figures 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}. The analysis of variance was performed with the following model:

@@

\section*{Additional Files}
  \subsection*{\textbf{Additional file 1}}
  Supplementary Figures (PDF 16 MB)
  \subsection*{\textbf{Additional file 2}}
  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
\end{backmatter}
\end{document}

>>>>

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:explanation}A; see also Supplementary methods). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines calling different methods) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments (or a desired subset), avoiding recomputing the same step twice without the need to save all alternative intermediates, and compiling evaluations (including run time) on the fly (see Supplementary methods for more detail). Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 2-3). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. In addition, we wanted to assess whether the effect of a given parameter alteration is robust to changes in the rest of the pipeline. We therefore monitored several complementary metrics across multiple steps of the process. We proceeded in a step-wise fashion, first testing a large variety of parameters at the early steps of the pipeline along with only a set of mainstream options downstream, then selecting main alternatives and proceeding to a more detailed benchmark of the next step (Figure \ref{fig:explanation}B).

@@

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figure 8-10). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that stood out across more than one dataset (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified{\color{red}, although such a pattern was not observed in the other datasets.}

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by the (real) subpopulations. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Pearson correlation >0.9, Supplementary Figure 19A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 19B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figure 20-21). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figure 20-22). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:selection}A and Supplementary Figure 20-21). \textit{Deviance} proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly performant) but did not perform so well to select genes with a high deviance explained.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figure 16 and 24). This tended to affect the accuracy of the clustering (Supplementary Figure 25), although in both cases (number of clusters and ARI), \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

Since ARI is dominated by differences in the number of clusters (Supplementary Figure 2-3) and no single metric is perfect, we diversified them (Figure \ref{fig:clust}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 3). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.

The MI score and minimum precision, which are largely independent of the estimated number of clusters, were overall higher for the \textit{walktrap} method (Figure \ref{fig:clust}), while the mean F1 score favored both \textit{scran} methods (\textit{walktrap} and \textit{fast greedy}) over \textit{Seurat}. Finally, the ARI score at the true number of clusters, when available, showed similar performances. However, because \textit{Seurat}'s resolution parameter had a large impact on the number of clusters identified (Supplementary Figure 2 and 24), \textit{Seurat} could always be coerced into producing the right number of clusters. Instead, the number of clusters found by \textit{scran} was considerably less influenced by the available parameters (number of nearest neighbors or steps in the random walk - see Supplementary Figure 28), and as a result \textit{scran}-based clustering sometimes never produced a partitioning with the right number of clusters. This observation, along with \textit{scran}'s higher MI score, suggest that \textit{scran} sometimes simply divides a real subpopulation into two clusters (possibly tracking some unknown biological differences) rather than committing misclassification errors. Overall, the \textit{walktrap} method appeared superior to the \textit{fast greedy} algorithm and was generally less prone to misclassification than \textit{Seurat} clustering, although the latter offered more control over the resolution. A major difference between \textit{walktrap}-based clustering and \textit{Seurat} is the computing time (Figure \ref{fig:clust}B). We found that using the Annoy approximation to the nearest neighbors somewhat reduced computing time at no apparent cost to accuracy. It nevertheless remained considerably slower than \textit{Seurat}. For all methods, some poorly distinguishable subpopulations from both the Zhengmix8eq and simMix1 datasets remained very inaccurately classified in regard to all metrics.

@@

To illustrate the applicability of \textit{pipeComp} to a completely different problem, we benchmarked the impact of surrogate variable analysis (SVA)-based methods on bulk RNAseq differential expression analysis (DEA). The rationale behind such approaches is that the identification of factors determining genome-wide variation (i.e. surrogate variables) and their inclusion in the DEA's generalized linear model might improve DEA, especially in the presence of sources of variation unrelated to the experimental variables. \textit{sva} \citep{leekCapturingHeterogeneityGene2007,leekSvaPackageRemoving2012} and \textit{RUVseq} \citep{rissoNormalizationRNAseq2014a} are two popular packages each implementing different methods to infer these variables. We therefore created a \texttt{PipelineDefinition} with three steps: 1) feature filtering, 2) removal of unwanted variation (i.e. SVA), and 3) DEA using the most established methods  \citep{germainRNAontheBENCHComputationalEmpirical2016}. The package's `pipeComp\_dea' vignette details the creation process. We developed 3 benchmark datasets: 1) the \texttt{ipsc} data is a random selection of 10 vs 10 samples from \citep{carcamoOriveAnalysisTranscriptional2017}, which are very heterogeneous, and foldchanges were manually added to 300 genes; 2) the \texttt{seqc} dataset contains 5 vs 5 samples of two mixtures of the SEQC consortium with different ERCC spike-in mixes \citep{ComprehensiveAssessment2014}, and includes a partially correlated batch effect; 3) A \texttt{simulation} of 8 vs 8 samples with two technical vectors of variation (see Methods for more details). As previously reported \citep{ComprehensiveAssessment2014}, we found that more stringent filtering improved DEA (Supplementary Figure 33A) and that the different DEA methods had similar performances (Figure \ref{fig:svadea}A and Supplementary Figure 33A). Including a SVA-based step increased sensitivity across all datasets, particularly in datasets with technical vectors of variation (\textit{seqc} and \textit{simulation}), and benefited all DEA methods in a similar fashion (Figure \ref{fig:svadea}A,C and Supplementary Figure 33-34). With the exception of the \texttt{RUVr} method, which led to a high False Discovery Rate (FDR), all other methods showed comparable gains in sensitivity while maintaining a decent error control (if a little above the nominal FDR), even when increasing the number of surrogate variables used (Figure \ref{fig:svadea}B-C). While the FDR did increase in the \textit{ipsc} dataset, we suspect that much of the effect could be due to spurious differences \citep{germainTamingHumanGenetic2017} between the groups which we wrongly assume to be false positives. Nevertheless, the FDR remained acceptable with up to three variables. These results therefore suggest that these methods are robust and can be widely applied.

@@

The set of alternatives found for the overall run (Figure 8 and Supplementary Figure 29-31) can be found in the \href{https://github.com/markrobinsonuzh/scRNA_pipelines_paper/blob/master/data/overall_run.Rmd}{paper's repository}. The analysis of variance was performed with the following model:

@@

\section*{Additional Files}
  \subsection*{\textbf{Additional file 1}}
  Supplementary Figure (PDF 16 MB)
  \subsection*{\textbf{Additional file 2}}
  More detailed description of \textit{pipeComp} and comparison with other tools. (PDF 0.5 MB)
\end{backmatter}
\end{document}
